---
title: "Endangered Languages"
format: html
execute: 
  echo: false
---


## How many are Languages are Endangered? 

According to the World Economic Forum there are approximately 7,000 known living languages. One-third of these languages are African and most of the world speaks one of 20 languages. Two-fifths of the world’s languages are positioned to die out.

Not all endangered languages are at equal risk of dying out. Here is UNESCO’s classification system which shows how vulnerable a language is:

| Classification | Explanation |
|----------------|-------------|
| Vulnerable     | Most children speak the language, but it may be restricted to certain domains                           (e.g., home). |
| Definitely Endangered | Children no longer learn the language as a 'mother tongue' in the home. |                
| Severely Endangered | Language is spoken by grandparents and older generations; while the parent                               generation may understand it, they do not speak it to children or among                                  themselves. |         
| Critically Endangered | The youngest speakers are grandparents and older, and they speak the language                            partially and infrequently. |            
| Extinct        | There are no speakers left. |


## Global Map Sphere 

The map below shows most countries scale on which level of endangered language their in based on the UNESCO’s classification system. 1 meaning on average it has vulnerable
languages all the way to 5 meaning on average it has extinct languages.  


```{r}
#| label: globe data
#| message: false
# Packages

# Data Manipulation
library(tidyverse) # For data manipulation
library(janitor)   # For clean column names, and other utilities
library(glue)      # String manipulation

# Interactive Maps
library(htmltools)
library(plotly)    # For interactive plots
library(leaflet)   # For interactive maps

# Map packages
library(sf)        # For map shape files
library(tigris)   
library(rnaturalearth)
library(ggspatial)

# Plotting packages
library(scales)
library(patchwork) 
library(RColorBrewer)
require(plotly)

# source('data_01_clean_languages_final.R')
# Below are the output if you run the line above
combined_data <- readRDS("combined_data.Rds")
country_doe_orig <- readRDS("country_doe_orig.Rds")
country_doe <- readRDS("country_doe.Rds")
covariates <- readRDS("covariates.Rds")
getMapGepProp <- readRDS("getMapGepProp.Rds")
```



```{r}
#| label: globe plot
#| message: false
#| column: screen

plot_geo(locations=country_doe_orig$country_codes,
         marker = list(colorbar = list(title = 'Country Median DoE')),
         colorscale="YlOrRd",
         text=paste(country_doe_orig$country_name, country_doe_orig$language_count, sep = "-"),
         z=country_doe_orig$degree_of_endangerment_median) |>
  layout(
    showlegend = T, geo = getMapGepProp(),
    title = 'World Language Degree of Endangerment Scale by Country'
  )
```


## World Map of the Classification of Endangered Langueages 

The map below shows the classification of endangerment of languages in each country that are know of so far.

```{r}
#| label: flat map data
#| message: false
# source('data_01_clean_flat_map_data.R')
# Below are the output if you run the line above
map_data <- readRDS("map_data.Rds")
```



```{r}
#| label: flat map plot
#| column: screen
#| message: false

color_pal <- colorFactor(
  palette = c(
    'green',
    'orange',
    '#D73377',
    '#4A0D67', 
    'red',
    '#6F0C17',
    'black',
    '#002642',
    'grey'
    
  ),
  domain = map_data$endangerment_degree
)


interactive_map <- leaflet() |>
  addTiles()

interactive_map <- interactive_map |> 
  addCircleMarkers(
    data = map_data,  # Use your data frame
    lng = ~longitude,  # Column name for longitude
    lat = ~latitude,  # Column name for latitude
    color = ~color_pal(endangerment_degree),  # Column name for the categorical variable
    radius = ~confidence_in_degree,  # Size of the markers
    stroke = FALSE,  # Remove stroke around markers
    fillOpacity = 0.7,  # Opacity of the markers
    label = ~name_in_the_language #endangerment_degree  # Display the category label on hover
  )



tag.map.title <- tags$style(HTML("
  .leaflet-control.map-title { 
    transform: translate(-50%,20%);
    position: fixed !important;
    left: 50%;
    text-align: center;
    padding-left: 10px; 
    padding-right: 10px; 
    background: rgba(255,255,255,0.75);
    font-weight: bold;
    font-size: 28px;
  }
"))

title <- tags$div(
  tag.map.title, HTML("Classification of Endangered Languages across the World.<br>Size of radius represent degree of belief in classification")
)  



interactive_map |>
  addLegend(data = map_data,
            "topright", pal = color_pal, values = ~endangerment_degree,
            title = "Degree of Endangerment",
            opacity = 1
  ) |>
  addControl(title, position = "topleft", className="map-title")
```

