---
title: "Endangered Languages"
format: html
execute: 
  echo: false
---


## How many are Languages are Endangered? 
(overview of what you did...)
Reported by the World Economic Forum there are approximately 7,000 recognized languages worldwide, 
African languages make up one-third of the total, while the majority of people globally communicate 
in only 20 languages. The following data was pulled from data sets from mrbalikci github endangered-languages. Mapping data was then used to visualize most countries scale on which level of endangered language their in as well as the classification of endangerment of languages in each country that are know of so far. I then estimated a Bayesian linear model with degree of endangerment as the dependent variable and literacy, infant mortality, agriculture, and net migration as an independent variable. The analysis found that 40% of the world's languages are in danger of disappearing completely.

It's important to note that not all endangered languages face an equal risk of extinction. UNESCO has devised a classification system that assesses the vulnerability of languages, providing insights into their level of endangerment.

| Classification | Explanation |
|----------------|-------------|
| Vulnerable     |The language is commonly spoken by children, but its usage might be limited to specific areas or contexts. (ex: home, etc) |
| Definitely Endangered | Children no longer acquire the language as their first language within their family environment. |                
| Severely Endangered | The language is spoken by grandparents and older generations, and although parents may comprehend it, they do not actively use it when communicating with their children or among themselves. |         
| Critically Endangered | The language is spoken partially and infrequently by the youngest speakers, who are typically grandparents or older individuals. |            
| Extinct        | There are no remaining individuals who can speak the language.|


## Global Map Sphere 

The map below shows most countries scale on which level of endangered language their in based on the UNESCO’s classification system. 1 meaning on average it has vulnerable
languages all the way to 5 meaning on average it has extinct languages.  


```{r}
#| label: globe data
#| message: false
# Packages

# Data Manipulation
library(tidyverse) # For data manipulation
library(janitor)   # For clean column names, and other utilities
library(glue)      # String manipulation

# Interactive Maps
library(htmltools)
library(plotly)    # For interactive plots
library(leaflet)   # For interactive maps

# Map packages
library(sf)        # For map shape files
library(tigris)   
library(rnaturalearth)
library(ggspatial)

# Plotting packages
library(scales)
library(patchwork) 
library(RColorBrewer)
require(plotly)

# source('data_01_clean_languages_final.R')
# Below are the output if you run the line above
combined_data <- readRDS("combined_data.Rds")
country_doe_orig <- readRDS("country_doe_orig.Rds")
country_doe <- readRDS("country_doe.Rds")
covariates <- readRDS("covariates.Rds")
getMapGepProp <- readRDS("getMapGepProp.Rds")
```



```{r}
#| label: globe plot
#| message: false
#| column: screen

plot_geo(locations=country_doe_orig$country_codes,
         marker = list(colorbar = list(title = 'Country Median DoE')),
         colorscale="YlOrRd",
         text=paste(country_doe_orig$country_name, country_doe_orig$language_count, sep = "-"),
         z=country_doe_orig$degree_of_endangerment_median) |>
  layout(
    showlegend = T, geo = getMapGepProp(),
    title = 'World Language Degree of Endangerment Scale by Country'
  )
```


## World Map of the Classification of Endangered Langueages 

The map below shows the classification of endangerment of languages in each country that are know of so far.

```{r}
#| label: flat map data
#| message: false
# source('data_01_clean_flat_map_data.R')
# Below are the output if you run the line above
map_data <- readRDS("map_data.Rds")
```



```{r}
#| label: flat map plot
#| column: screen
#| message: false

color_pal <- colorFactor(
  palette = c(
    'green',
    'orange',
    '#D73377',
    '#4A0D67', 
    'red',
    '#6F0C17',
    'black',
    '#002642',
    'grey'
    
  ),
  domain = map_data$endangerment_degree
)


interactive_map <- leaflet() |>
  addTiles()

interactive_map <- interactive_map |> 
  addCircleMarkers(
    data = map_data,  # Use your data frame
    lng = ~longitude,  # Column name for longitude
    lat = ~latitude,  # Column name for latitude
    color = ~color_pal(endangerment_degree),  # Column name for the categorical variable
    radius = ~confidence_in_degree,  # Size of the markers
    stroke = FALSE,  # Remove stroke around markers
    fillOpacity = 0.7,  # Opacity of the markers
    label = ~name_in_the_language #endangerment_degree  # Display the category label on hover
  )



tag.map.title <- tags$style(HTML("
  .leaflet-control.map-title { 
    transform: translate(-50%,20%);
    position: fixed !important;
    left: 50%;
    text-align: center;
    padding-left: 10px; 
    padding-right: 10px; 
    background: rgba(255,255,255,0.75);
    font-weight: bold;
    font-size: 28px;
  }
"))

title <- tags$div(
  tag.map.title, HTML("Classification of Endangered Languages across the World.<br>Size of radius represent degree of belief in classification")
)  



interactive_map |>
  addLegend(data = map_data,
            "topright", pal = color_pal, values = ~endangerment_degree,
            title = "Degree of Endangerment",
            opacity = 1
  ) |>
  addControl(title, position = "topleft", className="map-title")
```

